This is an example of article title

### 2019-06-01 paper/paper.tex
1st rev on the way...

\section{SIMD sorting techniques}
\label{sec:methods}
%
\chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library. Additional work may be required from the user selecting among the functions and also setting some numeric parameters. The {\tt chipsort!} function does not require any configuration and offers a great performance in a wide range of cases. The remaining functions are intended to be used with arrays of different sizes, and \verb chipsort_large  offers a good overview of all the techniques implemented in the module. Some of these techniques might also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose. \chipsort does not currently support data records with separate key and payload, what might allow significant optimizations. Only the sorting of simple data types such as 32 and 64-bit integers and floating-point numbers has been studied so far.

This section will present how \verb chipsort_large  works, detailing each step of the process. It mostly follows~\cite{DBLP:journals/pvldb/InoueT15} with a few modifications. This function achieves sorting by first splitting the input array in chunks that are sorted separately with \verb chipsort_medium!  and then merged simultaneously. This inner sort is performed in multiple steps:
\begin{enumerate}
\item sorting small blocks of data with a sorting network;
\item reordering vectors in-place according to a matrix transpose;
\item vectorized Combsort with a limited number of iterations;
\item regular insertion sort until the whole data is sorted.
\end{enumerate}

The remainder of this section will detail the multiple techniques employed in this process.

@@

A sorting network can be represented by a Knuth graph such as Figure~\ref{fig:sorting-network}. The graph represents a network that takes an array of four values $(A_1, A_2, A_3, A_4)$ and outputs the sorted sequence $(S_1, S_2, S_3, S_4)$. Each vertical line in the graph represents a comparator.
\begin{figure}[htb]
\centerline{\includegraphics[width=0.6\linewidth]{fig/sorting-network-4.pdf}}
\caption{A sorting network for 4 elements. The dashed lines delimit each step where operations are independent of each other.}
\label{fig:sorting-network}
\end{figure}

In \chipsort, as in the related work, parallelism is mainly exploited by carrying out the comparisons of the sorting network over vectors of size $V$ contained in SIMD registers, effectively running the network $V$ times simultaneously. Other forms of instruction-level parallelism can also be in play, though. In especial, it may also be possible to parallelize instructions such as \verb min  and \verb max , and also the loading of data into registers. This is often done implicitly by the microarchitecture, and programmers can only try to make sure their code is suitable for this by properly ordering operations and avoiding branching, for instance. Part of this work is expected to be performed by the compiler, and especially LLVM in the case of Julia.

\chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}. The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} works by assigning each element of the sequence at each step to a variable, and calculating values from the next step according to the network specification. The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types. Figure~\ref{fig:sort-net-listing}

@@

This function is a good first example of the utilization of Julia's meta-programming features in \chipsort. Multiple implementations of sorting networks are available~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,sortingnetworksjl,ultrasort}, however they implement networks of different sizes straight as code. In \chipsort they are represented as data structures that guide a generated function that produces equivalent code. While this data structure currently is hard-coded, it can potentially be produced only when necessary by an implementation of an algorithm such as Bose-Nelson, what remains a future plan for the module.

The {\tt sort\_net} generated function expects elements with {\tt max} and {\tt min} methods defined. It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works. In practice \chipsort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}. This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers.

\subsection{SIMD vectors transpose}
%
Consider a group of stacked SIMD vectors. Each column in this representation is termed a {\em lane}. The result from {\tt sort\_vecs} is that each lane contains a sorted sequence. This data must be transposed to produce sorted vectors. This is done in \chipsort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2. The generated code consists mostly of calls to {\tt SIMD.shufflevec}. This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions, contemplating each different case, and in \chipsort there is only a single higher-level Julia generated function that produces equivalent code.

Figure~\ref{fig:transpose-vecs} displays an example of 64 random integers loaded into SIMD registers as 4 vectors. After {\tt sort\_vecs} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers). The \verb chipsort_small!  function is essentially a sorting network followed by a transpose and a merge procedure.

@@

\subsection{Bitonic Merge Networks}
%
A bitonic sequence starts either increasing or decreasing and contains at most one change of direction. Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, {\tt bitonic\_merge}. Merging two sorted vectors requires first reversing one of them, and the result is stored in two new vectors of the same size containing the first and second halves of the total sorted sequence.

@@

\subsection{Vectorized Combsort}
%
One of the most peculiar aspects of \chipsort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm. Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376,INCERPI198737}. The array is swept multiple times, and pairs of values at a given distance from each other are compared and stored in order. The distance is reduced at each pass by a factor of $\frac{4}{3}$. This algorithm lends itself easily to vectorization: performing the comparisons over vectors of size $V$ effectively allows us to run the algorithm over $V$ lists simultaneously.

\chipsort contains two functions that utilize Combsort. The first is {\em chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the compiler optimizations. {\tt chipsort!} starts with Combsort until the interval size is 1, when it switches to Insertion sort. This is similar to what is often done in practical implementations of Quicksort, as in the Julia standard library, switching to Insertion sort when the array becomes too small. The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort is applied only once through the whole array. Shell sort also terminates naturally applying Insertion sort to a partially sorted array.

The other function is {\tt chipsort\_medium!}, an explicitly vectorized implementation utilizing {\tt SIMD.jl}. This second function also performs a number of other operations before and after using Combsort. It consists of the following steps:
\begin{enumerate}
\item Apply a sorting network on $K$ blocks of $J$ vectors of size $V$.
\item Vectorized Combsort until the interval size is 1.
\item Transpose blocks.
\item Transpose data in-place from $K$ blocks of $J$ vectors into $J$ blocks of $K$ vectors.
\item Vectorized Combsort again.
\item Sort blocks again.
\item Insertion sort.
\end{enumerate}

@@

Finishing a sorting method with Insertion sort is a common practice, present at least in Musser's Introsort~\cite{musser1997introspective}. The utilization of Insertion sort as a final stage to Combsort has been considered before in the literature~\cite{combwiki,INCERPI198737}, although this method does not seem to enjoy wide adoption. The main contribution of the present work may be to propose again this method in the SIMD context, extending AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, bringing experimental evidence of its good performance.

\subsection{In-place matrix transpose}
%
In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}. While carrying out this procedure is simple, it is not trivial to find out which are the cycles to tranpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[Sec. 1.3.3, Ex. 12]{DBLP:books/lib/Knuth97}.

\chipsort contains the generated function {\tt transpose!} to perform in-place matrix transposition. The cycle seeds are computed at the time of code generation, and at runtime the function only needs to move the data and compute the indices detecting when each cycle finishes. The function is therefore completely generic, supporting any dimensions, but with a very simple runtime code, with the most complex part of the problem being tackled programatically during code generation.

@@

The intermediate data from the tree is kept by \chipsort in a contiguous array, treated as a heap. Each node ancestor can be found simply by the Euclidean division of its index by 2. During operation it is assumed this data should fit some level of cache memory, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.

\section{Experiments}
\label{sec:experiments}
%
This section reports experiments with the methods offered by the \chipsort library, as well as alternative baseline methods. The term {\it method} can be understood as both the sorting algorithm and their actual implementations in the Julia language. The specific methods tested are:
\begin{itemize}
\item {\tt chipsort!} Combsort finishing with Insertion sort, written as a serial program and vectorized by the compiler.
\item {\tt chipsort\_small!} in-register, based on sorting networks.
\item {\tt chipsort\_medium!} vectorized Combsort, plus additional steps, finishing with Insertion sort.
\item {\tt chipsort\_large} merge-sort with a multi-way merge tree, starting from chunks processed by {\tt chipsort\_medium!}.
\item {\tt insertion\_sort!} Insertion sort taken from the Julia library.
\item {\tt sort!} the Julia standard sort, Quicksort finishing with Insertion sort for small arrays (20 elements).
\end{itemize}

@@

A simple experiment was also carried out with other programming languages in the same machine to provide a richer context. Table~\ref{tab:bench-cxx} diplays statistics collected from Julia, Python and C++ programs. For the case of Python the value obtained when calling the function through PyCall was better than measuring in the native Python shell, therefore the best value was kept. Results obtained for C++ with Julia interopt using Cxx.jl were slightly worse than native, thus discarded. While Numpy was almost as fast as Julia, no reason could be found for the significantly worse performance from the C++ standard libraries. Supposedly there are important optimizations missing.

>>>>

\section{SIMD sorting techniques}
\label{sec:methods}
%
\chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library. The {\tt chipsort!} function does not require any configuration and offers a great performance on a wide range of cases. The remaining functions are intended to arrays of different sizes, and require choosing some numeric parameters. The function \verb chipsort_large  offers a good overview of all the techniques implemented in the module, and this section is structured according to its stages. Some of these techniques might also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.

The way \chipsort  works mostly follows~\cite{DBLP:journals/pvldb/InoueT15}, although \chipsort does not support yet registers with key and payload. Another main difference is that Combsort is always employed with an Insertion sort finalization.

The function \verb chipsort_large  achieves sorting by first splitting the input array in chunks that are sorted separately with \verb chipsort_medium!  and then merged simultaneously. This inner sort is performed in multiple steps:
\begin{enumerate}
\item sorting small blocks of data with a sorting network;
\item reordering vectors in-place according to a matrix transpose;
\item vectorized Combsort with a limited number of iterations;
\item regular insertion sort until the whole data is sorted.
\end{enumerate}

The remainder of this section details the multiple techniques employed in this process.

@@

A sorting network can be represented by a Knuth graph such as Figure~\ref{fig:sorting-network}. Each vertical line is a comparator. The graph represents a network that takes an array of four values $(A_1, A_2, A_3, A_4)$ and outputs the sorted sequence $(S_1, S_2, S_3, S_4)$.
\begin{figure}[htb]
\centerline{\includegraphics[width=0.6\linewidth]{fig/sorting-network-4.pdf}}
\caption{A sorting network for 4 elements. The dashed lines delimit each step where operations are independent of each other.}
\label{fig:sorting-network}
\end{figure}

In \chipsort, as in the related work, parallelism is mainly exploited by carrying out the comparisons of the sorting network over vectors of size $V$ contained in SIMD registers, effectively running $V$ networks simultaneously. Other forms of instruction-level parallelism can also be in play, though. In especial, it may be possible to parallelize instructions such as \verb min  and \verb max , or the loading of data into registers. This is often done implicitly by the microarchitecture, and programmers can only make sure their code is suitable for this by properly ordering operations and avoiding branching, for instance. Part of this work is expected to be performed by the compiler, and especially LLVM in the case of Julia.

\chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}. The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} works by assigning each element of the sequence at each step to a variable, and calculating values from the next step according to the network specification. The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types. Figure~\ref{fig:sort-net-listing} shows one example of generated code.

@@

This function is a good first example of how \chipsort utilizes Julia's metaprogramming features. Multiple implementations of sorting networks are available in the literature~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,sortingnetworksjl,ultrasort}, however they implement networks of different sizes straight as code. In \chipsort they are represented as data structures that guide a generated function to produce equivalent code. While this data structure is currently hard-coded, it could be potentially produced only when necessary by an algorithm such as Bose-Nelson, what remains a future plan for the module.

The {\tt sort\_net} generated function expects elements with {\tt max} and {\tt min} methods defined. It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works. In practice \chipsort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}. This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers, or spilling data to the stack if all registers are occupied.

\subsection{SIMD vectors transpose}
%
Consider a group of stacked SIMD vectors. Each column in this representation is termed a {\em lane}. The result from {\tt sort\_vecs} is that each lane contains a sorted sequence. This data must be transposed to produce sorted vectors. This is done in \chipsort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2. The generated code consists mostly of calls to {\tt SIMD.shufflevec}. This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions contemplating each different case, while in \chipsort there is only a single Julia generated function that produces equivalent code.

Figure~\ref{fig:transpose-vecs} displays an example of 16 random integers loaded into SIMD registers as 4 vectors. After {\tt sort\_vecs} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers). The \verb chipsort_small!  function is essentially a sorting network followed by a transpose and a merge procedure.

@@

\subsection{Bitonic Merge Networks}
%
A bitonic sequence starts either increasing or decreasing and contains at most one change of direction. Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/bitonic-merge-network.jl#L12}{\tt bitonic\_merge}. Merging two sorted vectors requires first reversing one of them, and the result is stored into two new vectors of the same size containing the first and second halves of the total sorted sequence.

@@

\subsection{Vectorized Combsort}
%
One of the most peculiar aspects of \chipsort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm. Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{INCERPI198737,dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}. The array is swept multiple times, and pairs of values at a given distance from each other are compared and stored in order. The distance is reduced at each pass by a factor of $\frac{4}{3}$. This algorithm lends itself easily to vectorization, either by parallelizing the operations from a single array or executing the algorithm over independent arrays in parallel.

\chipsort contains two functions that utilize Combsort, employing parallelism in these two distinct forms. The first is {\tt chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the optimizing compiler. {\tt chipsort!} starts with Combsort until the interval size is 1, when it switches to Insertion sort. This is similar to what is often done in practical implementations of Quicksort, as in the Julia standard library, switching to Insertion sort when the array becomes too small. The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort is applied only once through the whole array.

The other function is {\tt chipsort\_medium!}, an explicitly vectorized implementation utilizing {\tt SIMD.jl}. This second function also performs a number of operations before and after Combsort. It consists of the following steps:
\begin{enumerate}
\item Apply a sorting network on $K$ blocks of $J$ vectors of size $V$.
\item Vectorized Combsort until the interval size is 1.
\item Transpose blocks.
\item Transpose data in-place from $K$ blocks of $J$ vectors into $J$ blocks of $K$ vectors.
\item Vectorized Combsort again.
\item Sort blocks again.
\item Insertion sort.
\end{enumerate}

@@

Finishing a sorting method with Insertion sort is a common practice, present at least in Musser's Introsort~\cite{musser1997introspective}. The utilization of Insertion sort as a final stage to Combsort has been considered before~\cite{combwiki,INCERPI198737}, although the method does not seem to enjoy wide adoption. The main contribution of the present work may be to propose it again in the SIMD context, extending AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and bringing experimental evidence of its good performance.

\subsection{In-place matrix transpose}
%
In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}. While carrying out this procedure is simple, it is not trivial to find out what are the cycles to tranpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[Sec. 1.3.3, Ex. 12]{DBLP:books/lib/Knuth97}.

\chipsort contains the generated function \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/transpose-vecs.jl#L19}{\tt transpose!} to perform in-place matrix transposition. The cycle seeds are computed at the time of code generation, and at runtime the function only needs to move the data and compute the indices detecting when each cycle finishes. The function is therefore completely generic, supporting any dimensions, but with a very simple runtime code, with the most complex part of the problem being tackled programatically during code generation.

@@

The intermediate data from the tree is kept by \chipsort in a contiguous array, treated as a heap. Assuming 1-based indexing, each node ancestor can be found simply by the Euclidean division of its index by 2. During operation it is assumed this data should fit some level of cache memory, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.

\section{Experiments}
\label{sec:experiments}
%
This section reports experiments with the methods offered by the \chipsort library, as well as alternative baseline methods. The term {\it method} can be understood as both the algorithms and their implementations in the Julia language. The specific methods tested are:
\begin{itemize}
\item {\tt chipsort!} Combsort finishing with Insertion sort, written as a serial program and vectorized by the compiler.
\item {\tt chipsort\_small!} in-register, based on sorting networks.
\item {\tt chipsort\_medium!} vectorized Combsort, plus additional steps, finishing with Insertion sort.
\item {\tt chipsort\_large} merge-sort with a multi-way merge tree, starting from chunks processed by {\tt chipsort\_medium!}.
\item {\tt insertion\_sort!} Insertion sort taken from the Julia library.
\item {\tt sort!} the Julia standard sort, Quicksort finishing with Insertion sort for small arrays (20 elements).
\end{itemize}

@@

\subsection{Other languages}
A simple experiment was also carried out with other programming languages in the same machine to provide a richer context. Table~\ref{tab:bench-cxx} diplays statistics collected from Julia, Python and C++ programs. For the case of Python the value obtained when calling the function through PyCall was better than measuring in the native Python shell, therefore the best value was kept. Results obtained for C++ with Julia interoperation using Cxx.jl were slightly worse than native, thus discarded. While Numpy was almost as fast as Julia, no reason could be found for the significantly worse performance from the C++ standard libraries. Supposedly there are important optimizations missing.
